---
title: "Identifying Schools with Biggest Differences in Math Test Scores, Part 2"
author: "Laura Charleston"
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

### Introduction

We'll do what we did in class, but with a twist: instead of Montgomery County, Pa., you'll be working with data from Montgomery County, Md. Same basic idea, though: linear regression looking at free and reduced-price meals and math test scores.

Just as in the exercise, you will create a scatterplot to visualize the relationship between the FARMS (Free and Reduced-price Meal System) percentage and the percentage of students who met or exceeded math expectations. You'll then use linear regression to identify the schools that show the biggest differences between their actual and expected math test scores and evaluate the results of your model.

### Loading the Data

First, let's load the data into a data frame using the read_csv() function from the tidyverse library.

```{r}
schools <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/montgomery_md.csv")
```
### Task: Create a Scatterplot and Describe the Results (5 points)

First, let's create a scatterplot to visualize the relationship between the FARMS percentage and the percentage of students who met or exceeded math expectations. We'll use ggplot2 to create the plot and add a linear regression line.

```{r}
schools |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations", title = "Scatterplot of Math Expectations vs FARMS Percentage")
```
Based on what you see here, what's the relationship between FARMS percentage and math test scores? Compare this scatterplot to the one you made in the class activity.

ANSWER HERE

There is a weak negative relationship between FARMS percentage and the math test scores, but the correlation is more defined than the scatterplot in the class activity. The scatterplot in the class activity had outliers too but the dots were not in a linear format. This scatterplot is more linear.

### Task: Linear Regression Model (7 points)

Next, you'll perform a linear regression analysis using the lm() function. We'll set the percentage of students who met or exceeded math expectations as the dependent variable, and the FARMS percentage as the independent variable. In other words, if we know the FARMS percentage for a school, how reliably can we predict its math score?

```{r}
model <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools)

summary(model)
```

Provide the model's output for the three statistics mentioned in the class activity.

ANSWER HERE

The model shows a significant negative relationship between FARMS percentage and math performance. The adjusted r-squared in 0.6734, meaning the model explains about 67% of the variation in scores. The coefficient for FARMS percentage is -0.59511, indicating scores drop as FARMS percentage increases. This relationship is statistically significant with a p-value of 0.00000000000000022. 


Based on what you've learned so far, what does it tell us about the nature of the relationship between FARMS percentage and math scores? Is this model worth exploring further? Why or why not?

ANSWER HERE

The model shows that as FARMS percentage goes up, the math scores go down. This means there is a negative relationship between poverty and math performance. The model explains a large part of the difference in scores, so it gives us useful information. Yes it is worth exploring further because it helps us understand how poverty may affect student achievement. 



### Task: Identifying Schools with Biggest Residuals (3 points)

Let's calculate and examine the residuals next. 

```{r}
schools$predicted <- predict(model)

# Calculate the residuals
schools <- schools |> 
  mutate(residual = `% Met or Exceeded Math Expectations` - predicted)

# Sort the data frame by the absolute value of the residuals in descending order
schools <- schools |>
  mutate(abs_residual = abs(residual)) |>
  arrange(desc(abs_residual))

```

Look at the absolute values of the residuals; what are the largest differences between the predicted result and the actual math scores? Pick two schools with the largest residuals. Find the zip codes where they are located and using the CensusReporter.org website, describe the most interesting demographic data from each zip code.

ANSWER HERE


1. Wayside Elementary 
- Has a high math expectancy,81%, and low FARMS percentage, 6% 
- According to CensusReporter.org, 1% of the children within this zipcode live below the poverty line. 
2.Travilah Elementary
- Has a high math expectancy, 81%, and low FARMS percentage, 6%
- According to CensusReporter.org, 7% of the children within this zipcode live below the poverty line. 



## Task: Evaluate the model (5 points)

Describe what conclusions you'd draw from this model's results and how it would influence a story pitch about the school board member and math scores.

ANSWER HERE

I would write a story about the correlation between schools that have positive residuals and the zipcode's poverty rate for children. Since FARM is used as a way to gauge the poverty rate of a school, I'd like to compare the FARMS percentage to the poverty rate of children.

I would also like to write a story about the education attainment. Just because a school has a negative stigma due to their math expectancy, does not mean they cannot/do not attain a higher level of education. 







