---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2024.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}

library(tidyverse)


```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}

wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv")

```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

ANSWER HERE

What stands out to me are the, in total, 11 advisers, directors, and one policy fellow that recorded 0 as their salary.  The highest salary is over 250,000 and lowest, disregarding the 0's, is 2,000. 

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}

mean(wh_salaries$salary)
  
median(wh_salaries$salary)

```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}


wh_salaries <- wh_salaries |>
  mutate( 
    min(wh_salaries$salary), 
    max(wh_salaries$salary), 
    mean_salary = mean(wh_salaries$salary),
    median_salary = median(wh_salaries$salary),
    )

```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

ANSWER HERE

The mean tells me the average of the salaries, which is 107,306. Still what stands out most for me is the 0 as the salary for some positions. I am curious why it is zero rather than n/a. The maximum tells me the highest paid position is an Associate Counsel, but there are others in the same position paid less. I am curious as what kind of associate counsel are they a part of and what is the difference. 


### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

ANSWER HERE

1) Answering why there are zeros. Maybe they were unpaid internship positions or something similar?
2) Identify the different positions and rank the highest to lowest paid positions. 
3) Using Associate Counsel as an example, find a position with varying salaries and investigate why. Could there be different departments with the same position?
4) Identify every person's sex and see who is paid more between men or women. Or which sex is in higher paying positions. 
