---
title: "HW4 MD CityCrime"
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}

library(tidyverse)

library(janitor)


```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}

md_cities <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv")


```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}


md_cities_rates <- md_cities |>
  mutate( violent_rate_2019 = violent2019/pop2019 * 10000 , property_rate_2019 = property2019/pop2019 * 10000 , violent_Rate_2020 = violent2020/pop2020 * 10000, property_rate_2020 = property2020/pop2020 * 10000)

```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?

ANSWER HERE

Violent crime rate in 2019
- 37 cities have a rate of 0% for violent crimes 
- highest violent crime rate was in Baltimore with 183%
- 4 cities with over 100% in violent crime rate was Baltimore (183%), Ocean City (131%), Williamsport (129%), and Elkton (681%). 

Property Rate in 2019
- 19 cities have 0% property crime rate
- Ocean city has over 1000% in property violence rate
- Top 3 cities with the highest property violence rates are Ocean City (1036%), Oakland (732%), and Elkton (681%)

Violent Crime rate in 2020
- 43 cities with 0% violent crime rates in 2020. Meaning 6 more cities recieved a 0% in violent crimes. 
- The highest violent crime rate is 239% in Ocean City. Baltimore dropped down to second city with the most violent crime rate. So more cities become safer and Ocean city became increased in violent crime rate, passing 2019. Ocean city was second before. 
- Top 3 cities with the highest violent crimes is Ocean City (239%), Baltimore (159%), and Cambridge (156%). 

Property Rate in 2020
- 20 cities have 0% property violence rate. One city more than last year. 
- Highest property violence rate is 1212%, Ocean City. 176% more than last year. Continues to be the city with the highest property violence rate. 

### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`. new-old/old

```{r}

md_cities_change <- md_cities |>
  mutate( violent_pct_chg =  ((violent2020 - violent2019)/violent2019) * 100 , property_pct_chg = ((property2020 - property2019)/property2019 ) * 100)


```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

ANSWER HERE

Violent 
- 12 cities had a 100% drop in violent crimes 
- 6 cities say Inf
- Highest violent crime increase was 900%, Edmonston 
- 9 cities had 0% change

Property
- 5 cities have a 100% drop in property crimes
- 4 cities say Inf 
- Highest property violence increase was 318%, Crisfield
- 4 cities have 0% change. 

### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

ANSWER HERE

2020 reports state Ocean City, Maryland tops Baltimore as the city with the most property and violent crimes. 

According to datasaet of total crimes reported by Maryland cities in 2019 and 2020. Baltimore city had the highest violent crime rate with 183%. Ocean city was second place in violent crimes but first in property crimes with 1036%. But in 2020, Ocean city's violent crime rate was 239% and property crime rate is 1212%. 

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
