---
title: "HW7_MandMs"
---

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse

```{r}

library(tidyverse)


```



### Task 2: Load the M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv


```{r}

m_and_m <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")


```

### Task 3: Sample 5 rows of data from the dataset

```{r}


sample_data <- sample_n(m_and_m, 5)



```



### Task 4: Calculate the mean age and standard deviation of the sample, using either the red or black number of M&Ms


```{r}

sample_data|>
  summarize( mean = mean(red), sd = sd(red))



```


### Task 5: Calculate the mean age and standard deviation of the population, using the same measure as in Task 4

```{r}

m_and_m |>
  summarize (mean = mean(total), sd = sd(total))

```



### Task 6: Compare the sample and population statistics

Compare the two sets of means and standard deviations. Are they similar?

No they are not. The population, which I calculated the mean and sd of the total column in the m&m dataset, is higher than the sample mean and sd. The mean is significantly greater, while the sd is higher by 9.  

### Task 7: Describe the results

Write a few sentences about what you found in the data. What does this tell you about the M&Ms dataset and about the act of random sampling?

When you are random sampling, you are getting a more precise number. Which is why the sample data's results, for the mean, is significantly lower, and the sd for sample data is lower, but not significantly. I would say this is why random sampling is preferred for studying a large group. 


When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
